Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-09-20 manuscript/manuscript.Rmd
Update all datasets

```{r setup, include=FALSE}
library(magrittr)
library(ltm)

@@

<!-- ^[<https://osf.io/8r3d4/?view_only=5ad7bafcd16b4d4ba08bb28b0e2bd02d>] -->
<!-- Commenting this out for IACM -->

@@

Advisers were randomly assigned to one of two treatments. In the
"Baseline" treatment, they only observed the score of the advisee on the
mathematics quiz. In the "Expectation" treatment, they observed the
score as well as the advisee's guess for how many questions they
answered correctly. Note that since the outcome is determined only by
the past score on the quiz, the advisee's guess is immaterial to which
group they should compete against. Moreover, in neither treatment did they
receive any demographic information about their advisee. Participants
gave recommendations to 10 advisees, which unbeknownst to them were five
men and five women matched to have identical performance on the test.^[We made this decision to account for the possibility of gender differences in performance.]
They were informed that if their advice was shown to a participant who
returned for the follow-up survey, they would receive the identical
bonus as that participant. The survey then concluded with basic
demographic questions.

@@

ggplot(df_exp, aes(x = Performance, y = avg_hardGroup, color = Gender)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),se = FALSE) +
  geom_smooth(data = df_none, aes(x = Performance, y = avg_hardGroup),
              method = "glm", method.args = list(family = "binomial"), color = "grey", se = FALSE) +
  scale_color_manual(values = c("firebrick3",'dodgerblue3', "grey"),
                     labels = c("Female + Expectations", "Male + Expectations", "Expectation Hidden")) +
  scale_y_continuous(name = "P(Compete Against High Performance)",
                     labels = scales::percent,
                     expand = c(0, 0)) +
  scale_x_continuous(name = "Performance",
                     expand = c(0, 0),
                     breaks = seq(0, 10, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))
```

@@

<!-- David: If we plotted on the x-axis performance and on the y-axis the difference -->
<!-- between recommendations for men & women, this would look inverse-U shaped, right? -->
<!-- Might be a nice way to illustrate what you describe. -->

<!-- Amanda: I think our primary purpose here is still showing the existence of gender difference -->
<!-- The current plot shows discrepancy between two lines, allowing us to demonstrate gender gap -->

To examine the quality of advice, we computed the expected bonus
earnings for someone who followed the recommendations. For example, if an
adviser suggested competing against the High Performer group, we matched the
advisee against all 20 members of that group and determined how often their score
matched or exceeded that of the member. We then multiplied this number by the
respective bonus earnings (50 cents and 30 cents for the High and Low Performer 
group, respectively). To see if including expectations leads to worse advise
for men, we report a linear probability model with the experimental treatment
of the adviser, the gender of the advisee, and their interaction in Column 3
of Table \@ref(tab:study1regs). Displaying expectations led men to be advised
to compete against the High Performer group more often, and this advise turned
out to be bad: men receive worse advise than do women when expectations are 
displayed, but not in their absence.^[This analysis was not preregistered, and we note here that the reduction in expected earnings is small. However, it is interesting that expectations have a negative effect for men who underestimate their performance on average. One possibility is that advisers suggest the High Performer group more often than is optimal. We return to the possibility that advice is overall biased to be flattering in Study 2.]

```{r study1time3}

@@

To determine whether flattering advice is truly costly, however, we need
to examine the outcome of the advisees. In particular, they could ignore flattering
advice, recognizing it as such and thus failing to adhere to it. Align with our prediction, participants who were primed with high expectations earned less when their expectations were conveyed to the advisers, although this result is only directional  (see Column 1 of Table \@ref(tab:study1time3)). Similarly, as shown in Column 2 of Table \@ref(tab:study1time3), we also found that male participants received less reward when their expectations were presented, even when their performance was the same. These findings suggest that flattering advice is not without consequences. Furthermore, our research indicates that because males are more likely to follow such advice (see Column 4 of Table \@ref(tab:study1time3), their tendency towards competition intensifies the cost of such advice.

## Discussion
When the quality of a decision is not dependent on someone's expectations,
advice should not be moved by whether an adviser is aware of the advisees
expectations. However, consistent with our account that advisers consider the
belief utility of the advisee and want to flatter, rather than disappoint them,
we find that displaying expectations to advisers does matter.
Specifically, we found that men underestimate their score on a mathematics test
less than do women. When these expectations are displayed to advisers, they
are more likely to tell men to compete against the High Performer group.
Notably, this is bad advice: men whose advisers observed their expectations
got worse advice as a result. Our findings suggest that males with the same scores as their female counterparts earned less, although the evidence is only directional. This may be due to males receiving more favorable advice and being more likely to adopt it, resulting in even worse outcomes.

<!-- David: One more sentence about the earnings difference here, which does not -->
<!-- differ significantly. And one sentence about the new analysis of whether -->
<!-- people follow the advice or not. -->

@@

<!-- David: Just commenting on what I tried to do here... a bit of a transition -->
<!-- for why this study follows from the last. -->
So far, we have shown that displaying expectations can induce gender
differences in advice-giving. Specifically, advisers were more likely to
recommend competing against a group of High Performers to men who had higher
expectations about their performance than did women. We now test directly whether this 
is driven by interpersonal concerns. Specifically, we examine whether 
more flattering advice is the result of a desire to be liked and whether
advisees indeed do like advisers more when they give flattering advice.

@@

ggplot(study2_time2, aes(x = Condition, y = Advice, fill = Condition)) +
    geom_violin(trim = FALSE) +
     scale_fill_manual(values = c("firebrick3",'dodgerblue3'), labels = c("Accurate", "Likable")) +
  geom_hline(yintercept = 7, col = 'grey', linetype = 'dashed', linewidth = 0.7) +
  scale_y_continuous(name = "Advised Rank", breaks = seq(1, 10)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

Next, we turn our attention to the advisers (Figure \@ref(fig:study2advice)). 
In the Accuracy condition, those
uploading selfies were advised to bet on rank `r round(mean(study2_time2[study2_time2$Condition == "Accurate",]$Advice),2)`.
Notably, this is significantly more attractive than the 7th rank those advisers
had themselves guessed just on the prior screen
(`r papaja::apa_print(t.test(study2_time2[study2_time2$Condition == "Accurate",]$Advice, mu = 7, var.equal = T))$statistic`).
This suggests that even when incentivized for accuracy, participants offered
flattering advice.^[This shading could be due to concerns of avoiding disappointment. However, it could also be that advisers are uncertain about the rankings they had given and make a recommendation that combines their own belief with a uniform prior.]
Importantly, and as predicted, we find that advisers in the Likeable treatment
recommend betting on a lower rank, communicating that they think the participant
in the selfie is more attractive
(`r round(mean(study2_time2[study2_time2$Condition == "Likable",]$Advice),2)`,
`r papaja::apa_print(t.test(Advice ~ Condition, data = study2_time2, var.equal = T))$statistic`).
This suggests that advisers inferred that advising someone that they
were more attractive would make the adviser appear more likeable and
therefore provided advice that communicated a more favorable impression
of the participants' attractiveness. The distribution shown in Figure \@ref(fig:study2advice)
shows that participants do not simply tell participants that they are the
most attractive person in the group. They may infer that flattering advise needs
to be somewhat realistic to be believable. We return to this in the general discussion.

<!-- David: Making a note here that we should be sure to discuss this at the end -->
<!-- as something for future research. -->

<!-- David: We should also report something on accuracy of the advice given across -->
<!-- the treatments. Even if no difference. -->

@@

<!-- Amanda: How about this -->
<!-- For warmth it looks interesting, but I think the plot is hard -->
<!-- to read. Let's hold off on this for now. -->

@@

<!-- The Figure is not super clear. How about we try a series of box plots, -->
<!-- with the x-axis as the advised rank? Standard errors will be huge at the -->
<!-- extremes. -->

<!-- Also, we need to do something relative to the advisees' expectations. -->
<!-- I wonder what would be interesting here? Maybe averages for three groups: -->
<!-- more attractive than the participants' own guess, equally attractive, and less -->
<!-- attractive? -->

<!-- Amanda: How about the following picture? -->
```{r study2rating, fig.cap="Participants tend to rate advisors who suggest betting on a higher rank (implying greater attractiveness) as more likable and warm."}

@@

# Create the first plot
plot1 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = likable),position = position_dodge(width = 0.8), alpha = 0.8, color = "black", fill = "firebrick3") +
  scale_y_continuous(name = "Likability") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

# Create the second plot
plot2 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = warmAvg),position = position_dodge(width = 0.8), alpha = 0.8, color = "black", fill = "dodgerblue3") +
  scale_y_continuous(name = "Warmth") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

ggplot(rating_df, aes(x = Advice, y = Freq, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("firebrick3",'dodgerblue3')) +
  labs(x = "Advice", y = "Rating") +
  ggtitle('Response to Advice') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20)) +
  geom_text(aes(label = paste0(Freq), y = 0.5), position = position_dodge(width = 0.9), size = 5, color = 'white', fontface = "bold")

@@

## Discussion
In a context where advice communicates ego-relevant information (here, people's
attractiveness), we find that the advice people give is contingent on their
incentives. Specifically, when they get rewarded for being more likeable, they
recommend that the advisee bet on a more favorable rank than when they are
incentivized for accuracy. Importantly, advisees do not discount flattering
advice and instead evaluate people who advise them to bet on a more attractive
rank as warmer and more likeable. These gains to interpersonal perceptions do
not come at the cost of trustworthiness, even when the advice is substantially
inflated (e.g., a recommendation that the advisee is the most attractive in the group).

# General Discussion
Advice has the potential to shape people's career and personal outcomes.
Honest feedback, however, may be painful to learn if it falls short of one's
expectations. As prior work notes, this may motivate people to avoid information 
and avoid seeking help [@Benabou2022; @Golman2017; @Jaroszewicz2022].
We present evidence from two experiments that advisers are also cognizant of this
cost. As a result, they present flattering advice that avoids disappointing the
recipient, and correctly anticipate that this boosts how advisees perceive them.
However, this flattering advice comes at a cost to advisees, who would do worse
if they followed it blindly, as we show in Study 1.

>>>>

```{r setup, include=FALSE}

library(magrittr)
library(ltm)

@@

<!-- ^[<https://osf.io/8r3d4/?view_only=5ad7bafcd16b4d4ba08bb28b0e2bd02d>] -->

<!-- Commenting this out for IACM -->

@@

Advisers were randomly assigned to one of two treatments. In the
"Baseline" treatment, they only observed the score of the advisee on the
mathematics quiz. In the "Expectation" treatment, they observed the
score as well as the advisee's guess for how many questions they
answered correctly. Note that since the outcome is determined only by
the past score on the quiz, the advisee's guess is immaterial to which
group they should compete against. Moreover, in neither treatment did
they receive any demographic information about their advisee.
Participants gave recommendations to 10 advisees, which unbeknownst to
them were five men and five women matched to have identical performance
on the test.[^1] They were informed that if their advice was shown to a
participant who returned for the follow-up survey, they would receive
the identical bonus as that participant. The survey then concluded with
basic demographic questions.

[^1]: We made this decision to account for the possibility of gender
    differences in performance.

@@

ggplot(df_exp, aes(x = Performance, y = avg_hardGroup, color = Gender)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),se = FALSE) +
  geom_smooth(data = df_none, aes(x = Performance, y = avg_hardGroup),
              method = "glm", method.args = list(family = "binomial"), color = "grey", se = FALSE) +
  scale_color_manual(values = c("#2A629A",'#FF7F3E', "grey"),
                     labels = c("Female + Expectations", "Male + Expectations", "Expectation Hidden")) +
  scale_y_continuous(name = "P(Compete Against High Performance)",
                     labels = scales::percent,
                     expand = c(0, 0)) +
  scale_x_continuous(name = "Performance",
                     expand = c(0, 0),
                     breaks = seq(0, 10, 2)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))
```

@@

<!-- David: If we plotted on the x-axis performance and on the y-axis the difference -->

<!-- between recommendations for men & women, this would look inverse-U shaped, right? -->

<!-- Might be a nice way to illustrate what you describe. -->

<!-- Amanda: I think our primary purpose here is still showing the existence of gender difference -->

<!-- The current plot shows discrepancy between two lines, allowing us to demonstrate gender gap -->

To examine the quality of advice, we computed the expected bonus
earnings for someone who followed the recommendations. For example, if
an adviser suggested competing against the High Performer group, we
matched the advisee against all 20 members of that group and determined
how often their score matched or exceeded that of the member. We then
multiplied this number by the respective bonus earnings (50 cents and 30
cents for the High and Low Performer group, respectively). To see if
including expectations leads to worse advise for men, we report a linear
probability model with the experimental treatment of the adviser, the
gender of the advisee, and their interaction in Column 3 of Table
\@ref(tab:study1regs). Displaying expectations led men to be advised to
compete against the High Performer group more often, and this advise
turned out to be bad: men receive worse advise than do women when
expectations are displayed, but not in their absence.[^2]

[^2]: This analysis was not preregistered, and we note here that the
    reduction in expected earnings is small. However, it is interesting
    that expectations have a negative effect for men who underestimate
    their performance on average. One possibility is that advisers
    suggest the High Performer group more often than is optimal. We
    return to the possibility that advice is overall biased to be
    flattering in Study 2.

```{r study3time3}

@@

To determine whether flattering advice is truly costly, however, we need
to examine the outcome of the advisees. In particular, they could ignore
flattering advice, recognizing it as such and thus failing to adhere to
it. Align with our prediction, participants who were primed with high
expectations earned less when their expectations were conveyed to the
advisers, although this result is only directional (see Column 1 of
Table \@ref(tab:study3time3)). Similarly, as shown in Column 2 of Table
\@ref(tab:study3time3), we also found that male participants received
less reward when their expectations were presented, even when their
performance was the same. These findings suggest that flattering advice
is not without consequences. Furthermore, our research indicates that
because males are more likely to follow such advice (see Column 4 of
Table \@ref(tab:study3time3), their tendency towards competition
intensifies the cost of such advice.

## Discussion

When the quality of a decision is not dependent on someone's
expectations, advice should not be moved by whether an adviser is aware
of the advisees expectations. However, consistent with our account that
advisers consider the belief utility of the advisee and want to flatter,
rather than disappoint them, we find that displaying expectations to
advisers does matter. Specifically, we found that men underestimate
their score on a mathematics test less than do women. When these
expectations are displayed to advisers, they are more likely to tell men
to compete against the High Performer group. Notably, this is bad
advice: men whose advisers observed their expectations got worse advice
as a result. Our findings suggest that males with the same scores as
their female counterparts earned less, although the evidence is only
directional. This may be due to males receiving more favorable advice
and being more likely to adopt it, resulting in even worse outcomes.

<!-- David: One more sentence about the earnings difference here, which does not -->

<!-- differ significantly. And one sentence about the new analysis of whether -->

<!-- people follow the advice or not. -->

@@

<!-- David: Just commenting on what I tried to do here... a bit of a transition -->

<!-- for why this study follows from the last. -->

So far, we have shown that displaying expectations can induce gender
differences in advice-giving. Specifically, advisers were more likely to
recommend competing against a group of High Performers to men who had
higher expectations about their performance than did women. We now test
directly whether this is driven by interpersonal concerns. Specifically,
we examine whether more flattering advice is the result of a desire to
be liked and whether advisees indeed do like advisers more when they
give flattering advice.

@@

ggplot(study2_time2, aes(x = Condition, y = Advice, fill = Condition)) +
    geom_violin(trim = FALSE) +
     scale_fill_manual(values = c("#2A629A",'#FF7F3E'), labels = c("Accurate", "Likable")) +
  geom_hline(yintercept = 7, col = 'grey', linetype = 'dashed', linewidth = 0.7) +
  scale_y_continuous(name = "Advised Rank", breaks = seq(1, 10)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

Next, we turn our attention to the advisers (Figure
\@ref(fig:study2advice)). In the Accuracy condition, those uploading
selfies were advised to bet on rank
`r round(mean(study2_time2[study2_time2$Condition == "Accurate",]$Advice),2)`.
Notably, this is significantly more attractive than the 7th rank those
advisers had themselves guessed just on the prior screen
(`r papaja::apa_print(t.test(study2_time2[study2_time2$Condition == "Accurate",]$Advice, mu = 7, var.equal = T))$statistic`).
This suggests that even when incentivized for accuracy, participants
offered flattering advice.[^3] Importantly, and as predicted, we find
that advisers in the Likeable treatment recommend betting on a lower
rank, communicating that they think the participant in the selfie is
more attractive
(`r round(mean(study2_time2[study2_time2$Condition == "Likable",]$Advice),2)`,
`r papaja::apa_print(t.test(Advice ~ Condition, data = study2_time2, var.equal = T))$statistic`).
This suggests that advisers inferred that advising someone that they
were more attractive would make the adviser appear more likeable and
therefore provided advice that communicated a more favorable impression
of the participants' attractiveness. The distribution shown in Figure
\@ref(fig:study2advice) shows that participants do not simply tell
participants that they are the most attractive person in the group. They
may infer that flattering advise needs to be somewhat realistic to be
believable. We return to this in the general discussion.

[^3]: This shading could be due to concerns of avoiding disappointment.
    However, it could also be that advisers are uncertain about the
    rankings they had given and make a recommendation that combines
    their own belief with a uniform prior.

<!-- David: Making a note here that we should be sure to discuss this at the end -->

<!-- as something for future research. -->

<!-- David: We should also report something on accuracy of the advice given across -->

<!-- the treatments. Even if no difference. -->

@@

<!-- Amanda: How about this -->

<!-- For warmth it looks interesting, but I think the plot is hard -->

<!-- to read. Let's hold off on this for now. -->

@@

<!-- The Figure is not super clear. How about we try a series of box plots, -->

<!-- with the x-axis as the advised rank? Standard errors will be huge at the -->

<!-- extremes. -->

<!-- Also, we need to do something relative to the advisees' expectations. -->

<!-- I wonder what would be interesting here? Maybe averages for three groups: -->

<!-- more attractive than the participants' own guess, equally attractive, and less -->

<!-- attractive? -->

<!-- Amanda: How about the following picture? -->

```{r study2rating, fig.cap="Participants tend to rate advisors who suggest betting on a higher rank (implying greater attractiveness) as more likable and warm."}

@@

# Create the first plot
plot1 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = likable),position = position_dodge(width = 0.8), alpha = 0.8, color = "darkgrey", fill = "#2A629A") +
  scale_y_continuous(name = "Likability") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

# Create the second plot
plot2 <- ggplot(study2_time3, aes(x = Advice)) +
  geom_boxplot(aes(y = warmAvg),position = position_dodge(width = 0.8), alpha = 0.8, color = "darkgrey", fill = "#FF7F3E") +
  scale_y_continuous(name = "Warmth") +
  scale_x_discrete(name = "Advised Rank") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20))

@@

ggplot(rating_df, aes(x = Advice, y = Freq, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("#2A629A",'#FF7F3E')) +
  labs(x = "Advice", y = "Rating") +
  ggtitle('Response to Advice') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.key.size = unit(1.2, "lines"),          
        legend.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 20)) +
  geom_text(aes(label = paste0(Freq), y = 0.5), position = position_dodge(width = 0.9), size = 5, color = 'white', fontface = "bold")

@@

## Discussion

In a context where advice communicates ego-relevant information (here,
people's attractiveness), we find that the advice people give is
contingent on their incentives. Specifically, when they get rewarded for
being more likeable, they recommend that the advisee bet on a more
favorable rank than when they are incentivized for accuracy.
Importantly, advisees do not discount flattering advice and instead
evaluate people who advise them to bet on a more attractive rank as
warmer and more likeable. These gains to interpersonal perceptions do
not come at the cost of trustworthiness, even when the advice is
substantially inflated (e.g., a recommendation that the advisee is the
most attractive in the group).

# General Discussion

Advice has the potential to shape people's career and personal outcomes.
Honest feedback, however, may be painful to learn if it falls short of
one's expectations. As prior work notes, this may motivate people to
avoid information and avoid seeking help [@Benabou2022; @Golman2017;
@Jaroszewicz2022]. We present evidence from two experiments that
advisers are also cognizant of this cost. As a result, they present
flattering advice that avoids disappointing the recipient, and correctly
anticipate that this boosts how advisees perceive them. However, this
flattering advice comes at a cost to advisees, who would do worse if
they followed it blindly, as we show in Study 1.
